---
title: "Vikings_EDA"
format: html
editor: visual
---

# Exploring Vikings NFL Data

```{r}
library(tidyverse)
vikes_data <-read_csv("final_vikings_data.csv")
```

```{r, include=FALSE}

## Getting to know the data
colnames(vikes_data)

```

```{r}
## Getting to know the data
dim(vikes_data)
vikes_data<-vikes_data|>
  filter(!is.na(posteam) & posteam=="MIN") # I'm going to get only the plays where the vikes are on offense
```

## Contingency Tables and Numeric Summaries

First, visualize vikings play type per down.

```{r}
table(vikes_data$down, vikes_data$play_type)
```

Seems like the Vikings are more likely to pass over run on later downs.

Let's look at counts and ratios of pass and run plays per year.

```{r}
library(lubridate)
vikes_table_1 <-vikes_data|>
  mutate(year = year(game_date)) |>
  group_by(year) |> 
  summarize(run_count=sum(play_type=="run",na.rm=TRUE ), pass_count = sum(play_type=="pass",na.rm=TRUE))|>
  mutate(run_ratio = run_count/(run_count+pass_count),pass_ratio = pass_count/(run_count+pass_count))

vikes_table_1
```

I want to create a table that shows average yards per play by year.

```{r}
vikes_table_2 <- vikes_data |>
  mutate(year = year(game_date)) |>
  group_by(year) |>
  summarize(
    avg_yards = mean(yards_gained, na.rm = TRUE),
    yards_sd  = sd(yards_gained, na.rm = TRUE)
  )
            
```

```{r}
vikes_table_2
```

Cool! Vikings were averaging a high 4.3 yards per play in 2009. The Vikings were 12-4.

Let's add the Vikings wins to this table to look at how yards per play relates to games won.

```{r}
# Adding in a vector with wins is easier than trying to extract this information from a play by play dataset.

wins<-c(12,6,3,10,5,7,11,8,13,8)

vikes_table_2$wins=wins

vikes_table_2
```

## Visualization

```{r}
library(ggplot2)
ggplot(vikes_table_2, aes(x=avg_yards, y = wins))+
  geom_point(color="purple")+
  geom_text(aes(label=year), hjust = 1.1, size = 3.5)+
  labs(
    title = "Average yards per play versus Wins",
    x = "Average Yards",
    y = "Wins"
  )+
  theme_minimal()+
   expand_limits(x = min(vikes_table_2$avg_yards) - 0.1)
```

There seems to be a loose positive trend between average yards per play and record.

Let's next look at yards/run plays and yards/pass plays by year.

```{r}
vikes_table_3 <- vikes_data |>
  mutate(year = year(game_date)) |>
  group_by(year) |>
  summarize(
    avg_yards = mean(yards_gained, na.rm = TRUE),
    yards_sd  = sd(yards_gained, na.rm = TRUE),
    avg_run_yrds = mean(ifelse(play_type=="run", yards_gained, NA), na.rm=TRUE),
    avg_pass_yrds = mean(ifelse(play_type=="pass", yards_gained, NA), na.rm=TRUE)
  )
vikes_table_3

# I want to visualize this so I'm going to pivot longer
vikes_table_3_long<-vikes_table_3|>
  rename(
    run = avg_run_yrds,
    pass = avg_pass_yrds
  )|>
  pivot_longer(cols =c(run,pass),
               names_to = "play_type",
               values_to= "average_yards")
vikes_table_3_long
```

```{r}
ggplot(vikes_table_3_long, aes(x = factor(year), y = average_yards, fill = play_type)) +
  geom_col(position = "dodge") +
  labs(
    title = "Average Yards per Play by Year",
    x = "Year",
    y = "Average Yards",
    fill = "Play Type"
  ) +
    scale_fill_manual(
    values = c("run" = "purple",   
               "pass" = "gold")  
  ) +
  theme_minimal()
```

-   Vikings fans will fondly remember 2012 as Adrian Peterson rushing for 2000+ yards in 2012 after tearing his ACL and winning MVP. SKOL. This year the Vikings had the highest rush yards/attempt of any year.

-   Vikings fans will also remember 2009 as the year Brett Favre threw for 4200 yards and took the Vikings to the NFC championship game, only to lose to the Saints after an infamous "too many men on the field" penalty, and "bountygate"\--a system that incentivized Saints defensive players to try to knock opposing players out of the game. This was the year with the highest pass yards/attempt.

-   Let us also not forget that Favre and Peterson have both endured their fair share of scandals, and I don't want to reminisce on their glory days without noting their complicated legacies.

Lets look at some other things:

```{r}
vikes_table_4 <- vikes_data |>
  mutate(year = year(game_date)) |>
  group_by(year) |>
  summarize(
    avg_yards = mean(yards_gained, na.rm = TRUE),
    yards_sd  = sd(yards_gained, na.rm = TRUE),
    avg_run_epa = mean(ifelse(play_type=="run", epa, NA), na.rm=TRUE),
    avg_pass_epa = mean(ifelse(play_type=="pass", epa, NA), na.rm=TRUE),
    avg_run_wpa = mean(ifelse(play_type=="run", wpa, NA), na.rm=TRUE),
    avg_pass_wpa = mean(ifelse(play_type=="pass", wpa, NA), na.rm=TRUE)
  )
vikes_table_4

# I want to visualize this so I'm going to pivot longer for EPA
vikes_table_4_long_ep<-vikes_table_4|>
  rename(
    run = avg_run_epa,
    pass = avg_pass_epa
  )|>
  pivot_longer(cols =c(run,pass),
               names_to = "play_type",
               values_to= "average_epa")
vikes_table_4_long_ep

## And pivot longer for WPA
vikes_table_4_long_wp<-vikes_table_4|>
  rename(
    run = avg_run_wpa,
    pass = avg_pass_wpa
  )|>
  pivot_longer(cols =c(run,pass),
               names_to = "play_type",
               values_to= "average_wpa")
vikes_table_4_long_wp

```

```{r}
ggplot(vikes_table_4_long_ep, aes(x = factor(year), y = average_epa, fill = play_type)) +
  geom_col(position = "dodge") +
  labs(
    title = "Average EPA per Play by Year",
    x = "Year",
    y = "Average EPA",
    fill = "Play Type"
  ) +
    scale_fill_manual(
    values = c("run" = "purple",   
               "pass" = "gold")  
  ) +
  theme_minimal()
```

-   Similar trends are visible here. Note that the 2010 Vikings were 6-10.

```{r}
ggplot(vikes_table_4_long_wp, aes(x = factor(year), y = average_wpa, fill = play_type)) +
  geom_col(position = "dodge") +
  labs(
    title = "Average WPA per Play by Year",
    x = "Year",
    y = "Average WPA",
    fill = "Play Type"
  ) +
    scale_fill_manual(
    values = c("run" = "purple",   
               "pass" = "gold")  
  ) +
  theme_minimal()
```

-   This is also a fascinating breakdown. Note that in 2009, running was not advantageous at all, although Adrian Peterson did run for 1300+ yards that year.

Let's try to visualize some other things. I want to try do a heatmap relating EPA and field position.\

```{r}
epa_by_field<-vikes_data|>
  group_by(yardline_100)|>
  summarize(avg_epa = mean(epa, na.rm=TRUE))
epa_by_field
```

```{r}
ggplot(epa_by_field, aes(x=yardline_100, y=1, fill = avg_epa))+
  geom_tile()+
  scale_fill_gradient2()+
  labs(
    title = "Vikings EPA by Field Position",
    x = "Yards from Opponent End Zone",
    y = "",
    fill = "Average EPA"
  ) 

```

I would like to facet this by play type.

```{r}

epa_by_field_type<-vikes_data|>
  filter(play_type %in% c("run","pass"))|>
  group_by(yardline_100,play_type)|>
  summarize(avg_epa = mean(epa, na.rm = TRUE), .groups = "drop")

ggplot(epa_by_field_type, aes(x=yardline_100, y=1, fill=avg_epa))+
       geom_tile()+
       scale_fill_gradient2()+
       facet_wrap(~play_type, ncol = 1) +
       labs(
    title = "Vikings EPA by Field Position and Play Type",
    x = "Yards from Opponent End Zone",
    y = "",
    fill = "Average EPA"
  )
         
  
```

Seems like passing generally has a higher EPA from just about anywhere in the field.
